---
title: "Exploration of BP44551 IL1Beta data"
format: html
toc: true
toc-depth: 3
toc-location: left
toc-expand: true
reference-location: margin
embed-resources: TRUE
params:
  password: 
---

```{r}
#| label: set-up
#| echo: FALSE

suppressPackageStartupMessages({
  library(tidyverse)
  library(knitr)
  library(kableExtra)
  library(mvtnorm)
  library(here)
  library(parallel)
  #library(rice)
  library(ggpubr)
  library(blandr)
  #library(irr)
  #library(psych)
  library(patchwork)
  library(plotly)
  #library(ggcompare)
})

#source("functions.R", echo = FALSE)
```

# Raw data of BP44551

```{r}
#| label: read dataset from OCEAN harbor
#| echo: false
#| eval: true

# read IL1B data second transfer
il1b <- arrow::read_parquet("/ocean/harbour/CDT71057/BP44551/CSRFinal_FA/ref/source/noncrf/EXL5_NC_TAI.parquet") |> 
  select(PATNUM, LABRESN, LABRESC, VISIT_CODE, VISIT_DATE, LABD, LBSTAT, LBLLOQ, LBULOQ) |> 
  mutate(VISIT_CODE=factor(VISIT_CODE, levels=c("PRN", "D1", "D15", "D29", "UNSCH"), ordered = TRUE)) |> 
  arrange(PATNUM, VISIT_CODE) 
  
   
protein <- arrow::read_parquet("/ocean/harbour/CDT71057/BP44551/CSRFinal_FA/ref/source/noncrf/EXL8_NC.parquet") |> 
  select(PATNUM, LABRESN, LABRESC, VISIT_CODE, VISIT_DATE, LABD, LBSTAT) |> 
  mutate(VISIT_CODE=factor(VISIT_CODE, levels=c("PRN", "D1", "D15", "D29", "UNSCH"), ordered = TRUE)) |> 
  arrange(PATNUM, VISIT_CODE)

# merge il1b and protein data by PATNUM and VISIT_CODE
dRaw <- il1b %>% 
  left_join(protein, by = c("PATNUM", "VISIT_CODE"), suffix = c("_IL1B", "_PROTEIN")) |> 
  mutate(AVAL_IL1B = case_when(
    LABRESC_IL1B == "BLQ" ~ LBLLOQ / 2,
    LABRESC_IL1B == "ALQ" ~ NA,
    TRUE ~ LABRESN_IL1B
  ))


```


Note: 
- The units are not the same, total protein is in ug/mL while the cytokines is in pg/mL. 
- Subject 10033 had a value ALQ on PRN visit, so it is removed from the analysis.
- 10021 and 10026 Unscheduled visit (mapped to PRN?); 10030 Unscheduled visit (mapped to D1?)

```{r }
#| label: data processing and check
#| echo: FALSE
#| warning: false

d <- dRaw |> 
  mutate(VISIT = case_when(
    PATNUM %in% c(10021, 10026) & VISIT_CODE == "UNSCH" ~ "PRN",
    PATNUM == 10030 & VISIT_CODE == "UNSCH" ~ "D1",
    TRUE ~ as.character(VISIT_CODE))
  ) |> 
  mutate(AVAL_IL1B_LOG = log(AVAL_IL1B), AVAL_PROTEIN = LABRESN_PROTEIN, AVAL_PROTEIN_LOG = log(LABRESN_PROTEIN)) |> 
  arrange(PATNUM, VISIT) |> 
  
  # divide IL1B by protein when protein is not NA
  mutate(RES = AVAL_IL1B / (AVAL_PROTEIN*1e6), logRES = log(RES))  
 
# Histogram
hist1 <- d %>% ggplot() + geom_histogram(aes(x=RES), bins = 6, fill="green", alpha=0.5) + xlab("IL1beta")
hist2 <- d %>% ggplot() + geom_histogram(aes(x=logRES), bins = 6, fill="red", alpha=0.5) + xlab("Log IL1beta")


# Boxplot
boxplot <- d %>% ggplot(aes(x=VISIT, y=logRES, color=VISIT)) +
  geom_boxplot() + 
#  stat_compare(aes(label = after_stat(sprintf("%s: %s", method, label))),
#               bracket = FALSE,
#               ymin = 140, 
#               ymax = 160,
#               ) +
  theme_bw()

# Spaghetti plot
spaghetti <- d %>% ggplot(aes(x=VISIT, y=AVAL_IL1B_LOG, label=PATNUM, color=as.factor(PATNUM))) + 
  geom_point(aes(group=PATNUM)) + 
  geom_line(aes(group=PATNUM)) 
  #geom_text(aes(label=PATNUM), color = "red", size = 3) 

(hist1 | hist2) 

(boxplot / spaghetti)

  
```



# Descriptive statistics

```{r}
#| label: summary
#| echo: FALSE

# mean and std
d %>% filter(VISIT %in% c("PRN", "D1",  "D15", "D29")) |> 
  group_by(VISIT) %>% 
  summarise(
    across(
      .cols = c(logRES, AVAL_IL1B_LOG),
      .fns = list(
        N = \(x) sum(!is.na(x)), 
        #Missing = \(x) sum(is.na(x)),
        #Mean = \(x) mean(x, na.rm = TRUE),
        SD = \(x) sd(x, na.rm = TRUE)
      )
    )
  )  %>% 
  kable(digits = 2)


```

# Correlation analysis

```{r}
#| label: correlation 
#| echo: false
#| warning: false

# transform dataset from long to wide format
d1 <- d %>% select(PATNUM, VISIT, RES, logRES) %>%
  #group_by(PATNUM, VISIT) %>% 
  #summarise(logRES = mean(logRES, na.rm = TRUE), .groups = "drop") %>% 
  pivot_wider(
    names_from = VISIT,
    values_from = c(RES, logRES)
  ) %>% 
  mutate(DIFF = RES_D1 - RES_PRN) 

d1 %>% select(-DIFF) |> kable(digits=1)


# raw scale
corp1 <- d1 %>% filter(!is.na(DIFF)) %>% 
  ggplot(aes(x = RES_PRN, y = RES_D1)) +
  geom_point() +
  #geom_text(aes(label=PATNUM), color = "red", size = 2) +
  stat_cor(p.accuracy = 0.001, r.accuracy = 0.01) +
  geom_smooth(method = "lm") +
  labs(x = "RUN-IN", y = "Day1", title = "Il1Beta") + 
  theme_classic()

# cor.test(d1$RES_PRN, d1$RES_D1, conf.level = 0.95, na.action="na.omit")

# log scale
# Calculate correlation and confidence interval
cor_test <- cor.test(d1$logRES_PRN, d1$logRES_D1, method = "pearson")
corr <- round(cor_test$estimate, 2)
corr_lci <- round(cor_test$conf.int[1], 2)
corr_uci <- round(cor_test$conf.int[2], 2)
conf_level <- attr(cor_test$conf.int, "conf.level") * 100

corp2 <- d1 %>% filter(!is.na(DIFF)) %>% 
  ggplot(aes(x = logRES_PRN, y = logRES_D1)) +
  geom_point() +
  #geom_text(aes(label=PATNUM), color = "red", size = 2) +
  stat_cor(p.accuracy = 0.001, r.accuracy = 0.01) +
  geom_smooth(method = "lm") +
  labs(x = "RUN-IN", y = "Day1", title = "log-Il1Beta") + 
  theme_classic() +
  annotate("text", 
           x = min(d1 %>% filter(!is.na(DIFF)) |> pull(logRES_D1)), 
           y = max(d1 %>% filter(!is.na(DIFF)) |> pull(logRES_PRN)),
           label = paste0("\n", conf_level, "% CI: [", corr_lci, ", ", corr_uci, "]"),
           hjust = 0, vjust = 1, size = 4,
           color = "black")

cowplot::plot_grid(corp1, corp2, labels = "auto", label_size = 12, align = "h")

```

# Agreement analysis

```{r}
#| label: ICC 
#| echo: false
#| eval: false

# icc calculation 
d2 <- d1 %>% filter(!is.na(DIFF))
ICC <- psych::ICC(cbind(d2$RES_PRN, d2$RES_D1))$results
#psych::ICC(cbind(d2$RES_PRN, d2$RES_D1))$lme # output variance component
kable(ICC %>% filter(type=="ICC3"), digits = 2) # type ICC3 is what we want

# reulsts cross validated using lmer or aov
d3 <- d %>% filter(PATNUM!=10016) 
#summary(lme4::lmer(RES ~ 1+ (1|PATNUM) + VISIT, d3))
#summary(lme4::lmer(RES ~ 1+ (1|PATNUM) + (1|VISIT), d3))

#summary(lme4::lmer(logRES ~ 1+ (1|PATNUM) + VISIT, d3))

#summary(lme4::lmer(logRES ~ 1+ (1|PATNUM) + (1|VISIT), d3))
#psych::ICC(cbind(d2$logRES_PRN, d2$logRES_D1))$lme # output variance component

# irr:icc seems only provide two-way random effects (ICC2) rather than two-way mixed effects (ICC3)
#irr::icc(cbind(d1$RES_PRN, d1$RES_D1), model = "twoway", type = "agreement", unit = "single")
```

## Bland-Altman analysis

Plot of differences between two measurements. Shaded areas present confidence interval limits for mean and agreement limits.

```{r}
#| label: Bland-Altman
#| echo: false
#| warning: false

# run the below one line code to generate plot  
# with(d2, blandr.draw(RES_PRN, RES_D1, annotate = TRUE))

d2 <- d1 %>% filter(!is.na(DIFF))

statistics.results <- with(d2, blandr.statistics(RES_PRN , RES_D1))
blandr.plot.ggplot( statistics.results,
                    plotTitle = "Bland-Altman plot for comparison of 2 measurements",) +
  geom_text(aes(label=round(statistics.results$bias,1)), x=200, y=0, color="white")
```

Plot of differences between two measurements (log-transformed). Shaded areas present confidence interval limits for mean and agreement limits.

```{r}
#| label: Bland-Altman analysis on log-scale
#| echo: false
#| warning: false

# run the below one line code to generate plot  
# with(d2, blandr.draw(RES_PRN, RES_D1, annotate = TRUE))

statistics.results.log <- with(d2, blandr.statistics(logRES_PRN , logRES_D1))
blandr.plot.ggplot( statistics.results.log,
                    plotTitle = "Bland-Altman plot for comparison of 2 measurements",) +
  geom_text(aes(label=round(statistics.results.log$bias,1)), x=4, y=0, color="white")
```


Plot of differences between two measurements, expressed as percentages of the values on the axis \[(PRN – D1)/((PRN + D1)/2)%)\], vs. the mean of the two measurements.

```{r}
#| label: Bland-Altman - plot difference as percentage
#| echo: false
#| warning: false
#| eval: true

d1 <- d1 %>%  
  mutate(RES_PRN_REL=2*RES_PRN/(RES_PRN+RES_D1)*100,
         RES_D1_REL= 2*RES_D1 /(RES_PRN+RES_D1)*100
  )
# Passes data to the blandr.statistics function to generate Bland-Altman statistics
statistics.results.rel <- with(d1, blandr.statistics( RES_PRN_REL , RES_D1_REL ))

# replace the relative means with raw means for plot
statistics.results.rel$means <-  statistics.results$means

# Generates a ggplot, with no optional arguments
blandr.plot.ggplot( statistics.results.rel,
                    plotTitle = "Bland-Altman plot for comparison of 2 measurements",) + 
  labs(y= "Difference/Average(%)") +
  geom_text(aes(label=round(statistics.results.rel$bias,1)), x=200, y=0, color="white")
```
