---
title: "Repeated-Measures / Within-Subjects ANOVA in R"
format: html
editor: visual
embed-resources: true
---

# **One-way repeated measures**

<https://www.statskingdom.com/repeated-anova-calculator.html>

<https://www.geeksforgeeks.org/repeated-measures-within-subjects-anova-in-r/>

We will use a simulated dataset of Repeated-Measures. In this example, we'll assume we have a dataset where subjects are measured at 2 different time points.

```{r}
library(tidyverse)
library(ggpubr)
library(corrplot)
library(mvtnorm)
```

## Step 1: Simulate Data

Treatment effect is defined as % mean change from baseline, assume data follows logNormal distribution. It's easy to know that $log(T2)=log(T1)+log(1-CfbPct/100)$

```{r}
# Set seed for reproducibility
set.seed(123)

# Simulate data for 10 subjects measured at 3 time points
CfbPct <- 70

subject <- factor(rep(1:100))
T1 <- 4 + rnorm(100, mean = 0, sd = 1)
#T2 <- T1 + log(1-CfbPct/100)  # No variablity at all
#T2 <- T1 + rnorm(100, mean = log(1-CfbPct/100),  sd=1) # No residual in this setting
T2 <- 4 + rnorm(100, mean = 0, sd = 1) + rnorm(100, mean = log(1-CfbPct/100),  sd=1) # Add residual, ie, variability that can not be explained by subject and time), but without considering the intra-subject correlation

# Create a data frame
data <- data.frame(subject, T1, T2)
```

### Check the correlation and variability

```{r}
data %>% 
  select(-subject) %>% ggplot() + 
  geom_point(aes(x=T1, y=T2), color="red") +   
  stat_cor(aes(x=T1, y=T2), p.accuracy = 0.001, r.accuracy = 0.01, color="red")  + 
  geom_point(aes(x=T1, y=T2_1), color="blue") +
  stat_cor(aes(x=T1, y=T2_1), p.accuracy = 0.001, r.accuracy = 0.01, 
           color="blue", label.x.npc="center")

out1 <- data %>% summarise(sdT1=sd(T1), 
                           sdT2=sd(T2), 
                           COV=cov(T1, T2))

out1 %>% mutate(r=COV/(sdT1*sdT2))
```

### Simulate data considering intra-subject correlation

$yij=a+𝞫1*treatment+𝞼(time)+𝞼$

```{r}
# Set seed for reproducibility
set.seed(123)

# Simulate data for 10 subjects measured at 3 time points
CfbPct <- 70

# generate data using multivariate normal distribution
nPerArm <- 3
nTot <- nPerArm*2
rho <- 0.7
sigma <- 1 # between subject
withinsub <- 0.5 # within subject

v <- matrix(c(sigma^2, rho * sigma^2, rho * sigma^2, sigma^2), ncol = 2)

sigma <- matrix(c(1, 0.7, 0.7, 1), ncol=2)

tibble(subject=rep(1:nTot, each=2),
               time=rep(1:2, nTot),
               arm=c(rep(0, nTot), rep(1, nTot))
       ) %>% 
  rowwise() %>% 
  mutate(rmvnorm(n=1, mean=c(4, 4+log(1-CfbPct/100)*arm), sigma=v) )
 
cor(response)
```

### Step 2: Fit the Repeated-Measures ANOVA Model

```{r}

dLong <- data %>% select(-T2_1) %>% pivot_longer(cols=c(T1,  T2),
                               names_to = "time", 
                               values_to = "response")

# Perform Repeated-Measures ANOVA
aov_model <- aov(response ~ time + Error(subject), data = dLong)

# View the summary of the ANOVA
summary(aov_model)
```

```{r}
lmer_model <- lme4::lmer(response ~ time+(1|subject), data=dLong, REML = TRUE)
summary(lmer_model)

```

```{r}
data %>% group_by(time) %>% summarise(mean(response))

data %>% summarise(mean(response))

# overall sum of square 
ss_tot <- var(data$response)*(30-1)

# sum of square for between time
ss_time <- 10*((55.36234-54.76448)^2+
    (53.07152	-54.76448)^2+
    (55.85958 -54.76448)^2
)

# sum of square for within subject
d1 <- data %>% group_by(subject) %>% summarise(subjmean=mean(response)) %>% bind_cols(
  data %>% pivot_wider(names_from = time, values_from = response) %>% 
  select(-subject) ) %>% 
  mutate(dev1=((T1-subjmean)^2+(T2-subjmean)^2+(T3-subjmean)^2),
         overallmean=mean(subjmean),
         dev2=(subjmean-overallmean)^2)

ss_sub_wthin <- d1 %>% summarise(sum(dev1))

# sum of square for between subject
ss_sub_between <- d1 %>% summarise(sum(dev2))
  


c(ss_tot, ss_time, ss_sub_wthin, ss_sub_between)
#611.937+112.0754+44.22794-948.1631

```
